---
title: '`r paste0(Sys.Date()," skeeter forecast")`'
output: pdf_document
---

```{r initiate, echo=FALSE, message=FALSE, warning=FALSE}
# set rmarkdown options
here::i_am("R/skeeter-forecast-template.Rmd")
knitr::opts_chunk$set(echo = FALSE, warning = FALSE, message = FALSE)
## get data
library(fitdistrplus)
library(forecast)
library(here)
library(kableExtra)
library(brms)
library(tidyverse)

#
covInt = function(center = NULL , int = NULL,...){
  # calculate the coverage interval around a value
  # this function returns two values:
  # lower = center-int
  # upper = center+int
  lower = center - int
  upper = center + int
  
  return(list(lower = lower,
              upper = upper))
}
## set options
options(mc.cores = parallel::detectCores())
rstan::rstan_options(auto_write = TRUE)
theme_set(theme_minimal())

## user setting
raft_filepath = sprintf("C:/Users/%s/OneDrive - UNT System/AERI Seed Insect Carbon/Data/2024_aeri_seed_mosquito_data.xlsx", Sys.info()[['user']])

pred_filepath = sprintf("C:/Users/%s/OneDrive - UNT System/Shared Documents - Leaf Litter Hot Moments/General/2024_predictions_aeri_seed_mosquito.xlsx", Sys.info()[['user']])
```

```{r data import}
## this script imports and tidys up raft count data
raft_data = readxl::read_excel(raft_filepath) %>%
  select(-Block, -Mesocosm, -Treatment, -Site)

# identify columns with numbers
raftNumCols = names(raft_data) %>% as.numeric %>% is.na %>% `!`

# rename cols to readible dates
raft_data = raft_data %>%
  rename_with(~as.character(openxlsx::convertToDate(.x, origin='1900-01-01')), .cols = names(.)[raftNumCols])

pred_data = readxl::read_excel(pred_filepath, skip = 1) %>%
  filter(is.na(name) |
         name != 'Total')
predNumCols = names(pred_data) %>% as.numeric %>% is.na %>% `!`

pred_data = pred_data %>%
  rename_with(~as.character(openxlsx::convertToDate(.x, origin='1900-01-01')), .cols = names(.)[predNumCols]) %>% 
  mutate(across(-c(name), as.numeric))

# count_vec = readxl::read_excel(pred_filepath)%>% filter(name == 'Total')
count_df = raft_data %>% select(which(raftNumCols)) %>% colSums(na.rm = FALSE) %>% enframe(name = 'date', value = 'count') %>% 
  mutate(date = as.Date(date))

current_date = as.character(Sys.Date())

name_df = c('GD' = 'Grace',
            'AM' = 'Arya',
            'EAC' = 'Emily',
            'JD' = 'Dr. D',
            'JMR' = 'Julia',
            'JRJ' = 'Jim',
            'JRB' = 'Dr. B',
            'ZC' = 'Z',
            'ARM' = 'Angel',
            'CRG' = 'Chris',
            'JK' = 'Juwan')


```

Good afternoon! Here is your Skeeter report for `r format(Sys.Date(), "%B %d, %Y")`. 

&nbsp;

Today we counted `r count_df %>% filter(date == eval(current_date)) %>% select(count) %>% unlist` skeeter rafts. The closest prediction today was made by 
`r name_df[which.min(abs(pred_data[[eval(current_date)]] - count_df %>% filter(date == eval(current_date)) %>% select(count) %>% unlist %>% unname))]` with a prediction of `r pred_data[which.min(abs(pred_data[[eval(current_date)]] - count_df %>% filter(date == eval(current_date)) %>% select(count) %>% unlist %>% unname)),eval(current_date)]`. Get your predictions in before tomorrow's count for a chance to show off your prediction skills!

```{r time series, fig.align='center', fig.height=4, fig.width=7, fig.cap = "Daily predictions (past = grey circles, current = black circles) and observations (black line) of mosquito rafts counted in mesocosms."}
# create a long plot data frame of predictions
pred_plotDf = pred_data %>% 
  pivot_longer(-name, names_to = 'date', values_to = 'prediction') %>% 
  mutate(date = as.Date(date),
         old = ifelse(as.Date(date)<(Sys.Date()+1), TRUE, FALSE))

# set plotting attributes
ylims = c(0,10^(signif(log10(max(c(count_df$count,na.omit(pred_plotDf$prediction)))))))

labelsDf = data.frame(date = as.numeric(count_df$date), datelab = as.character(format(count_df$date, "%b-%d")), day = paste0("Day ",1:nrow(count_df)))# %>% unite('label', date:day, sep = "") %>% unlist %>% unname

# do the plotting
ggplot()+
  geom_point(data = pred_plotDf, aes(x = as.numeric(date), y = prediction, color = old), shape = 21, fill = 'white', size = 1.2)+
  geom_line(data = count_df , aes(x = as.numeric(date), y = count ), linewidth = 1.1, color = 'black')+
  scale_y_continuous(name = "Skeeter Rafts (#)", limits = ylims)+
  scale_x_continuous(breaks = labelsDf$date, labels = labelsDf$datelab, sec.axis = dup_axis(labels = labelsDf$day))+
  scale_color_manual(values = c('black','darkgrey'))+
  theme(axis.title.x = element_blank(),
        legend.position = 'none',
        axis.text.x.bottom = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
        axis.text.x.top = element_text(size = 10, angle = 45, hjust = 0, vjust = 0),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())
  

```

To measure prediction accuracy, we will calculate the mean absolute error,

$$MAE = \frac{\sum_{i}^{n}|y_{i} - x_{i}|}{n}$$

where, $y_i$ is the prediction, $x_i$ is the observed value, and $n$ is the number of observations.

\newpage

### Rankings

```{r rankings table}
# create prediction accuracy table
pred_accDf = pred_plotDf %>% 
  left_join(count_df, by = 'date') %>% 
  mutate(diff = prediction - count) %>% 
  filter(!is.na(diff)) %>% 
  group_by(name) %>% 
  mutate(MAE = sum(abs(diff))/n()) %>% 
  select(name, date, diff, MAE) %>% 
  arrange(MAE) %>% 
  pivot_wider(id_cols = c(name, MAE), names_from = 'date', values_from = 'diff') %>%
  relocate(MAE, .after = last_col())
  
knitr::kable(pred_accDf, caption = "The rankings to date are:") %>% 
  kableExtra::kable_styling(latex_options = 'scale_down')
```

### Cumulative patterns

```{r cumulative patterns}

cumulative_df = count_df %>% 
  bind_rows(data.frame(date = as.Date("2024-06-26"),
                       count = 0),.) %>% 
  mutate(cumsum = cumsum(count))
 

ylims = c(0, ceiling(max(cumulative_df$cumsum, na.rm = TRUE)/10)*10)

labelsDf = data.frame(date = as.numeric(cumulative_df$date), datelab = as.character(format(cumulative_df$date, "%b-%d")), day = paste0("Day ",0:(nrow(cumulative_df)-1)))

cumulative_df %>% 
  ggplot()+
  geom_line(aes(x = date, y = cumsum), linewidth = 1.1)+
  geom_point(aes(x = date, y = cumsum), color = 'black', size = 1.1)+
  scale_y_continuous(name = "Cumulative Skeeter Rafts (#)", limits = ylims)+
  scale_x_continuous(breaks = labelsDf$date, labels = labelsDf$datelab, sec.axis = dup_axis(labels = labelsDf$day))+
  theme(axis.title.x = element_blank(),
        legend.position = 'none',
        axis.text.x.bottom = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
        axis.text.x.top = element_text(size = 10, angle = 45, hjust = 0, vjust = 0),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())

```

## Forecasts

Here are a few different models to forecast raft counts for `r Sys.Date()+1`. 

### Previous value

The simplest prediction is to simply predict the previous raft count. 

```{r last value}

lastVal = count_df %>% select(count) %>% unlist %>% is.na %>% `!` %>% which %>% max

ylims = c(0, ceiling(max(count_df$count, na.rm = TRUE)/5)*5)

count_df %>% 
  ggplot()+
  geom_line(data = count_df[c(lastVal, lastVal+1),] %>% fill(count,.direction = 'down'), aes(x = date, y = count), linewidth = 1.2, color = 'grey')+
  geom_point(data = count_df[c(lastVal, lastVal+1),] %>% fill(count,.direction = 'down') %>% slice_tail(n = 1), aes(x = date, y = count), size = 2, color = 'grey')+
  geom_line(aes(x = date, y = count), linewidth = 1.1, color = 'black')+
  geom_point(aes(x = date, y = count), size = 1.2, color = 'black')+
  scale_y_continuous(name = "Skeeter Rafts (#)", limits = ylims)+
  scale_x_continuous(breaks = labelsDf$date, labels = labelsDf$datelab, sec.axis = dup_axis(labels = labelsDf$day))+
  theme(axis.title.x = element_blank(),
        legend.position = 'none',
        axis.text.x.bottom = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
        axis.text.x.top = element_text(size = 10, angle = 45, hjust = 0, vjust = 0),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())
  

```


```{r last value accuracy}

lastVal_accDf = readRDS(here::here("data/lastVal_accDf.rds")) %>% 
  merge(count_df, by = 'date', all = TRUE) %>% 
  select(date, count = "count.y", prediction)

lastVal = count_df %>% ungroup %>% select(count) %>% unlist %>% is.na %>% `!` %>% which %>% max

lastVal_pred = 
count_df[c(lastVal, lastVal+1),] %>% fill(count,.direction = 'down') %>% slice_tail(n = 1) %>% rename(prediction = 'count')

lastVal_accDf = lastVal_accDf %>% merge(lastVal_pred, by = 'date', all = TRUE) %>% 
  mutate(prediction = coalesce(prediction.x, prediction.y)) %>% 
  select(-prediction.x, -prediction.y)

saveRDS(lastVal_accDf, here::here("data/lastVal_accDf.rds"))
  
  lastVal_accDf %>% 
  mutate(diff = prediction - count) %>% 
  filter(!is.na(diff)) %>% 
  mutate(MAE = sum(abs(diff))/n()) %>% 
  select(date, diff, MAE) %>% 
  arrange(MAE) %>% 
  pivot_wider(id_cols = MAE, names_from = 'date', values_from = 'diff') %>%
  relocate(MAE, .after = last_col()) %>% 
  knitr::kable(caption = "Mean predictive deviation of last value approach") %>% 
  kableExtra::kable_styling(latex_options = 'scale_down')

```

### Global Average

Another simple prediction is to use the global average. This method allows for a calculation of uncertainty based on the variation we observe over time. Importantly, day-to-day variability in egg raft numbers is not related to any process, but arises from random noise centered around some relatively fixed mean value.

```{r global average}

lastVal = count_df %>% select(count) %>% unlist %>% is.na %>% `!` %>% which %>% max

globAvg_accDf = readRDS(here::here("data/globAvg_accDf.rds")) %>% 
  merge(count_df, by = 'date', all = TRUE) %>% 
  select(date, count = "count.y", prediction, lower, upper)

globAvg = count_df %>% select(count) %>% unlist %>% mean(., na.rm = TRUE)

globAvg_int = count_df %>% select(count) %>% unlist %>% sd(na.rm = TRUE) %>% `/`(lastVal) %>% `*`(1.96) %>% covInt(center = globAvg, int = .)

# add newest prediction into the data.frame
globAvg_accDf[(lastVal+1),'prediction'] <- globAvg
globAvg_accDf[(lastVal+1), 'lower'] <- globAvg_int[[1]]
globAvg_accDf[(lastVal+1), 'upper'] <- globAvg_int[[2]]

#save the dataframe
saveRDS(globAvg_accDf, here::here("data/globAvg_accDf.rds"))

ylims = c(0, ceiling(max(c(count_df$count, globAvg_accDf$upper), na.rm = TRUE)/5)*5)

globAvg_accDf %>% 
  ggplot()+
  geom_line(data = globAvg_accDf[c(lastVal,lastVal+1),] %>% mutate(count = coalesce(count, prediction)), aes(x = date, y = count), linewidth = 1, color = 'grey')+
  geom_errorbar(data = globAvg_accDf [(lastVal+1),] , aes(x = date, ymin = lower, ymax = upper), linewidth = 1.2, color = 'grey', width = 0)+
  geom_point(data = globAvg_accDf [(lastVal+1),], aes(x = date, y = prediction), size = 2, color = 'grey')+
  geom_line(aes(x = date, y = count), linewidth = 1.1, color = 'black')+
  geom_point(aes(x = date, y = count), size = 1.2, color = 'black')+
  scale_y_continuous(name = "Skeeter Rafts (#)", limits = ylims)+
  scale_x_continuous(breaks = labelsDf$date, labels = labelsDf$datelab, sec.axis = dup_axis(labels = labelsDf$day))+
  theme(axis.title.x = element_blank(),
        legend.position = 'none',
        axis.text.x.bottom = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
        axis.text.x.top = element_text(size = 10, angle = 45, hjust = 0, vjust = 0),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())

```

```{r global average accuracy}

globAvg_accDf %>% 
  mutate(diff = prediction - count) %>% 
  filter(!is.na(diff)) %>% 
  mutate(MAE = sum(abs(diff))/n()) %>% 
  select(date, diff, MAE) %>% 
  arrange(MAE) %>% 
  pivot_wider(id_cols = MAE, names_from = 'date', values_from = 'diff') %>%
  relocate(MAE, .after = last_col()) %>% 
  knitr::kable(caption = "Mean predictive deviation of global average approach") %>% 
  kableExtra::kable_styling(latex_options = 'scale_down')

```

However, this approach ignores an important bit of information---the fact that egg rafts are counts and can only take whole numbers (i.e., 1,2,3,...).

## More complex predictions

We can start to make more complex predictions. The best way to begin this is to switch to making predictions at the mesocosm-level and scale up to total counts. This will allow us to possibly include more specific information to the experiment. As the global average example above highlighted, we have to think about the type of data we are taking, in this case counts. There are a number of distributions available for use with count data such as the Poisson and Negative Binomial distributions. Let's take a look at these distributions compared to the most recent distribution of counts.

```{r meso count hist, fig.cap = 'The distribution of egg raft counts from the most recent sampling (bars). We can see the difference in the predictions from the Poisson (blue line) and Negative Binomial (red line) distributions compared to the Guassian (black lines).'}

# build prediction from counts of previous day using poisson likelihood
last_counts = raft_data %>%
  dplyr::select(max(which(apply(raft_data, 2, function(x) all(!is.na(x))))))

colnames(last_counts) <- 'count'

norm_pars = fitdist(last_counts$count, 'norm', 'mle')[['estimate']]
pois_pars = fitdist(last_counts$count, 'pois', 'mle')[['estimate']]
negbin_pars = fitdist(last_counts$count, 'nbinom', 'mle')[['estimate']]

ggplot()+
  geom_density(data = data.frame(count = rnorm(1e3, mean = norm_pars['mean'], norm_pars['sd'])), aes(x = count, after_stat(ndensity)), linewidth = 1.1, color = 'black')+
  geom_density(data = data.frame(count = rpois(1e3, pois_pars['lambda'])), aes(x = count, after_stat(ndensity)), linewidth = 1.1, color = 'blue')+
  geom_density(data = data.frame(count = rnegbin(1e3, negbin_pars['size'], negbin_pars['mu'])), aes(x = count, after_stat(ndensity)), linewidth = 1.1, color = 'red')+
    geom_histogram(data = last_counts, aes(x = count, after_stat(ndensity)), alpha = 0.5)+
  scale_x_continuous(name = "Egg rafts (#)", limits = c(-0.2,NA), breaks = seq.int(0:6))+
  scale_y_continuous(name = "Relative frequency")+
  theme(axis.title.y = element_blank(),
        axis.text.y = element_blank())

```

### Poisson

```{r poisson preds}
#build a model to esimate total count from individual sample distribution
pois_simple = brm(count ~ 0 + Intercept,
                  data = last_counts,
                  family = poisson(),
                  prior = prior(normal(0, 10),
                                class = b,
                                coef = "Intercept"),
                  thin = 1,
                  iter = 1000,
                  file = here::here("data/models/pois_simple"),
                  file_refit = 'on_change',
                  chains = 4)

# convert to draws
pois_draws = as_draws_df(pois_simple, variable = 'b_Intercept')
# transform and sample from Poisson with posterior draws
pois_pred = map_dbl(unlist(pois_draws$b_Intercept), ~sum(rpois(27,exp(.x))))

```

```{r pois simple plot}

lastVal = count_df %>% select(count) %>% unlist %>% is.na %>% `!` %>% which %>% max

poisSimple_accDf = readRDS(here::here("data/poisSimple_accDf.rds")) %>% 
  merge(count_df, by = 'date', all = TRUE) %>% 
  select(date, count = "count.y", prediction, lower, upper)

poisSimpleMedian = pois_pred %>% median

poisSimpleInt = pois_pred %>% quantile(c(0.025, 0.975))

# add newest prediction into the data.frame
poisSimple_accDf[(lastVal+1),'prediction'] <- poisSimpleMedian
poisSimple_accDf[(lastVal+1), 'lower'] <- poisSimpleInt[1]
poisSimple_accDf[(lastVal+1), 'upper'] <- poisSimpleInt[2]

#save the dataframe
saveRDS(poisSimple_accDf, here::here("data/poisSimple_accDf.rds"))

ylims = c(0, ceiling(max(c(count_df$count, poisSimple_accDf$upper), na.rm = TRUE)/5)*5)

poisSimple_accDf %>% 
  ggplot()+
  geom_line(data = poisSimple_accDf[c(lastVal,lastVal+1),] %>% mutate(count = coalesce(count, prediction)), aes(x = date, y = count), linewidth = 1, color = 'grey')+
  geom_errorbar(data = poisSimple_accDf[(lastVal+1),] , aes(x = date, ymin = lower, ymax = upper), linewidth = 1.2, color = 'grey', width = 0)+
  geom_point(data = poisSimple_accDf[(lastVal+1),], aes(x = date, y = prediction), size = 2, color = 'grey')+
  geom_line(aes(x = date, y = count), linewidth = 1.1, color = 'black')+
  geom_point(aes(x = date, y = count), size = 1.2, color = 'black')+
  scale_y_continuous(name = "Skeeter Rafts (#)", limits = ylims)+
  scale_x_continuous(breaks = labelsDf$date, labels = labelsDf$datelab, sec.axis = dup_axis(labels = labelsDf$day))+
  theme(axis.title.x = element_blank(),
        legend.position = 'none',
        axis.text.x.bottom = element_text(size = 10, angle = 45, hjust = 1, vjust = 1),
        axis.text.x.top = element_text(size = 10, angle = 45, hjust = 0, vjust = 0),
        panel.grid.minor.y = element_blank(),
        panel.grid.minor.x = element_blank())

```

```{r pois simple accuracy}

poisSimple_accDf %>% 
  mutate(diff = prediction - count) %>% 
  filter(!is.na(diff)) %>% 
  mutate(MAE = sum(abs(diff))/n()) %>% 
  select(date, diff, MAE) %>% 
  arrange(MAE) %>% 
  pivot_wider(id_cols = MAE, names_from = 'date', values_from = 'diff') %>%
  relocate(MAE, .after = last_col()) %>% 
  knitr::kable(caption = "Mean predictive error of a simple Poisson model") %>% 
  kableExtra::kable_styling(latex_options = 'scale_down')

```

more forecasts to come???
